Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper-discussion, update

\Genetics2level{Superiority of single step models}
\subsection{Exp1}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the reduced set ($\Delta(n) = 54$) was not large enough to
provide considerably more useful information.
The latter was the case for \citeNP{Ashraf2016}, who imputed genotypes for about 
10,000 wheat lines using pedigree information and observed greater accuracy of
the single step method over genomic BLUP for all four evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations, modeled as a
fixed effect.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp2}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even minor gains, when using transcriptomic and genomic data in a
single step, might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
for the whole set of available parent lines rose considerably for several
agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
genomic information alone.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
This suggests that single step prediction should always be considered when the
incomplete predictor offers an appreciably higher predictive ability compared
to the complete predictor.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining to what extent individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in Exp1 we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20-30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable.

\section*{Conclusions}
We successfully applied single step prediction to a hybrid plant data set 
while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could elucidate the influence of differences
between two predictors on predictive ability when used in a single step
approach.
When the discrepancy in predictive ability between the complete and the
incomplete predictor was skewed profoundly in favor of the incomplete predictor,
such as for pedigree and transcriptomic data, single step prediction was shown
to be highly beneficial.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.

>>>>

\Genetics2level{Superiority of single step models}
\subsection{Exp1}
When using a single step approach based on complete genomic and incomplete
transcriptomic information, gains in predictive ability for the inbred lines
were at best just slightly higher than when using complete genomic information
alone.
One of the reasons might be that the subset of genotypes that was covered by
transcriptomic data was not a good representation of the full genetic space of
all available genotypes.
Another possibility is that the number of additional phenotypes in the full
set compared to the reduced set ($\Delta(n) = 54$) was not large enough to
provide considerably more useful information.
A larger dataset was available to~\citeNP{Ashraf2016}, who imputed genotypes
for about 10,000 wheat lines using pedigree information and observed greater
prediction accuracy of the single step method over genomic BLUP for all four
evaluated traits.
Predictive abilities reported for the maize diversity panel in this study were
slightly different from those reported by \citeNP{Guo2016}, for three
possible reasons:
1) We reduced the dataset to 164 tropical/subtropical lines with little evident
population structure and excluded all other genotypes from our analyses whereas
\citeNP{Guo2016} used genotypes from four different subpopulations with a model
using a fixed effect to account for the differences between subpopulations.
2) We applied quality checks for the predictor data after generating the subset
of 164 inbred lines, whereas \citeNP{Guo2016} applied the quality checks to 368
genotypes.
3) While \citeNP{Guo2016} used five-fold cross-validation with 500 repetitions,
we employed a leave-one-out cross-validation scheme (LOOCV).
Notwithstanding, the relative differences between predictive abilities for G 
and T, respectively, were similar in both studies.

\subsection{Exp2}
Hybrid breeding --- with noteworthy commercial applications in pig
\cite{Xiang2016,Tusell2016} and maize breeding~\cite{TheRoyalSociety2009} --- 
is a particularly challenging application field for prediction tools 
\cite{Kadam2016}.
Here, $2n$ parent individuals from two genetically distinct heterotic groups are
crossed to each other, yielding $n^{2}$ potential hybrid progeny that would
require intensive field testing.
In medium-sized plant breeding programs, the advent of the doubled haploid (DH) 
technology~\cite{Wedzony2009} now allows for an annual production of 1,000 
parent lines in each heterotic group, amounting to $10^{6}$ putative hybrid
progeny.
\citeNP{Westhues2017} have shown that, in hybrid breeding, the probability 
of successfully selecting the best observed genotypes based on the best 
predicted candidates is a strongly convex function of the predictive ability.
Thus, even minor gains, when using transcriptomic and genomic data in a
single step, might justify additional investments in RNA-seq at least for a
subset of genotypes that represents the genetic space of the breeding 
material well.
For the hybrid data, the predictive abilities of genomic information
increased considerably in the full set of genotypes compared to the reduced
set for several agronomic traits.
Hence, the inclusion of trancsriptomic data could not further improve upon
genomic information alone.
Predictive abilities obtained when using pedigree data could be 
improved considerably when, depending on the agronomic trait, combined with 
either transcriptomic or genomic information.
This suggests that single step prediction should always be considered when the
incomplete predictor offers an appreciably higher predictive ability compared
to the complete predictor.

\Genetics2level{Coverage of the genetic space}
Whereas animal breeding populations are oftentimes so large that the assembly 
of a core set of individuals with records on multiple predictors can effectively 
be done at random once more than 10,000 animals have pedigree records 
\cite{Fragomeni2015,Lourenco2015,Masuda2016}, pedigrees in plant breeding 
programs are typically considerably smaller.
Single step prediction offers the promise of leveraging the predictive ability
for all genotypes by borrowing the superior performance of an auxiliary
predictor on only a subset of genotypes, thereby reducing predictor costs.
Hence, we were interested in determining to what extent individuals should be 
complemented with information on another predictor.
By randomly and repeatedly declaring trancsriptomic information missing from
genotypes in Exp1 we examined how reliably single step prediction works
depending on the fraction of genotypes covered by both, the complete and the
incomplete predictor.
We observed that about $20-30\%$ of genotypes needed to be covered by both
predictors to match the performance achieved with the complete predictor
individually.
Beyond this fraction of covered genotypes the trend in predictive ability
changes was stable in either direction.
Predictive abilities increased simultaneously with $x$ when the incomplete
predictor outperformed the complete predictor in the reduced set of genotypes
and decreased with increasing $x$ when the complete predictor was the better
predictor.
This pattern clearly elucidates the influence on the difference between the
two predictors on predictive ability in a single step approach.

\section*{Conclusions}
We successfully applied single step prediction to an inbred and a hybrid plant
data sets while imputing with a quantitative downstream `omics' predictor.
By declaring different subsets of individuals as covered by one or two 
predictors, respectively, we could elucidate the influence of differences
between two predictors on predictive ability when used in a single step
approach.
When the discrepancy in predictive ability between the complete and the
incomplete predictor was skewed profoundly in favor of the incomplete predictor,
such as for pedigree and transcriptomic data, single step prediction was shown
to be highly beneficial.
Extensions to more than two predictors were outside the scope of this study
but, with mounting interest in systems genetics, should be considered in the
future.
